A comprehensive framework to capture the arcana of neuroimaging analysis

### 2019-03-20 manuscript/manuscript.tex
finished revising for resubmission

Study classes encapsulate a study dataset (i.e. data collected across
multiple subjects using the same acquisition protocol) with the suite of
analytical methods that can be applied to the dataset. The hierarchy of
a study dataset is assumed to have two levels, \emph{subjects} and
\emph{sessions}, with each session for each subject corresponding to a
specific \emph{visit}, e.g. timepoint in longitudinal study. Derivatives
can be created at any point in this hierarchy: per-session, per-subject,
per-visit and per-study. Iteration over subjects and visits is handled
implicitly by the framework. All Study classes must inherit from the
\emph{Study} base class and be created by the \emph{StudyMetaClass}
metaclass or subclasses thereof.

@@

At the heart of each Study class is the data specification attribute, which
specifies the input and output data of the analysis, and all
stages in between. There is a one-to-one relationship between entries in
the data specification and derivatives that are stored in the repository,
or will be stored if and when they are derived. Which intermediate
derivatives to include in the data specification, and therefore
store in the repository, is left to the discretion of the researcher
designing the analysis. However, as a general rule, derivatives that
require manual QC or are likely to be reused between different branches
of analysis should be included in the data specification.

Entries in the data specification can be of either a
\emph{file set} or \emph{field} type. File sets represent single files,
or sets of related files typically contained within a directory
(e.g. a multi-volume DICOM dataset). Fields
represent integer, floating point, date-time, or character string
values. By default, a field represents a single value but if the \emph{array}
flag is set, a field represents a list of values.
Each file set references a \emph{FileFormat} object, which
specifies the formats of the files in the set. File formats are
explicitly registered by the researcher at design time using the
\emph{FileFormat.register(format)} class method to avoid conflicts where
the same extension is used for different formats in different contexts.

The data specification of a Study class are defined by providing a
list of \emph{FilesetSpec} and \emph{FieldSpec} objects in the
\emph{add\_data\_specs} class attribute (Figure \ref{fig:data_spec}).
The lists are collated with data specifications in any base Study
classes during the creation of the class. Input data to the study are
distinguished from derivatives by using the \emph{FilesetInputSpec}
and \emph{FieldInputSpec} subtypes. However, the distinction between
input and derived data is somewhat blurred, as derivative specifications can be
passed input data when the class is instantiated. In addition,
when subclassing a Study class, or combining it in a MultiStudy class,
derivative specifications can be overridden by input
specifications, and vice-versa.

@@

All data specifications have a \emph{frequency} attribute which specifies
where the data sits in the hierarchy of the
dataset and can take the values 'per\_session',
'per\_subject, 'per\_visit' or 'per\_study'. In
addition, derived specifications are passed the name of a method in the
class that constructs the pipeline to derive them. Therefore, while a
pipeline can have multiple outputs, each derivative is derived by only
one pipeline.

@@

The syntax for pipeline construction is inspired by changes proposed for Nipype v2.0
(\url{github.com/nipy/nipype/issues/2539}).
Within a pipeline constructor method, a Pipeline object is constructed by the \emph{Study.pipeline} method
(Figure \ref{fig:pipeline_constructor}). At initialisation, each pipeline is assigned a name,
which must be unique amongst the pipelines constructed by the Study given the any combination of
switches. The methods implemented by the pipeline can be characterised by providing a the list of citations
and a text description to the \emph{citations} and \emph{desc} keyword arguments, respectively.

The \emph{add(name, interface)} method is used to add a node to a pipeline and takes
a unique name for the node (within the pipeline) and a Nipype Interface object, and returns
a reference to the newly added node. For clarity, is recommended to put all static inputs
(i.e. parameters) of the interface as keyword arguments of the interface constructor
(Figure \ref{fig:pipeline_constructor}). However, if an input conditionally depends on a parameter of the
study it can be set via the \emph{inputs} attribute of the node (e.g. \lstinline{my_node.inputs.my_param = 1.0}).

Node interfaces are connected to each other, and to inputs and outputs of the pipeline, by
providing \emph{inputs} and \emph{outputs} keyword arguments to the \emph{add} method. Both
arguments take a dictionary. The keys of the \emph{inputs} dictionary correspond to
trait names in the node's input interface, whereas the keys of the \emph{outputs} dictionary correspond to
names of entries in the data specification. The values of both dictionaries are 2-tuples. For pipeline
inputs, values of the \emph{inputs} dictionary consist of a name of an entry in the data specification
and the format the input data is expected in (i.e. a FileFormat for Fileset specifications or core Python type for Field specifications).
For pipeline outputs, values of the \emph{outputs} dictionary consist of a trait name in the node's output interface and the format the
output data is generated in. For input connections from other nodes, values of the \emph{inputs} dictionary
consist of a reference to the upstream node and the name of a trait in the
upstream node's output interface (output connections to other nodes are implied by input connections to the receiving nodes).

@@

Any external software packages required by a node should be referenced
in the \emph{requirements} keyword argument as a list of
\emph{Requirement} objects when the node added to the pipeline.
Similarly, the expected memory requirements in MB and
wall time for the node execution should be provided to the keyword arguments
\emph{memory}, and \emph{wall\_time}.

@@

Study parameters can be accessed during pipeline construction with the
\emph{Study.parameter(name)} method. If conditional logic is included in
the workflow construction that alters the pipeline inputs, outputs or
parameters then it should be controlled by a switch instead of a
parameter. The analysis branch designated by a switch value should be
tested with \emph{Study.branch(name)} in the case of boolean switches and
\emph{Study.branch(name, ref\_value)} in the case of string switches.

@@

Each Study instance is assigned a name, which is used to differentiate its
results from alternative analyses on the same dataset (e.g. with different
parameterisations). Parameters are set on initialisation of the Study object
along with the range of subject and visit IDs to be included in the analysis
If subject and visit IDs are not explicitly provided then all IDs found in the repository are included.
The remaining arguments passed to the Study object at initialisation are the
Repository, Processor and Environment modules and a list of
\emph{Input} objects to match data in the repository with inputs in the
data specification (Figure \ref{fig:study_application}).

@@

Data from group analyses (i.e. with \emph{per\_subject}, \emph{per\_visit}, and
\emph{per\_study} frequencies\emph{)} are stored in specially named
subjects and visits (e.g. `group'), the names for which are specified when
the repository is initialised. For example, given a basic
repository using `group' as the summary name for both subjects and visits,
per\_subject data for `subj1' would be stored at
\emph{\textless{}root\textgreater{}/subj1/}group, per\_visit data for
`visit1' in \emph{\textless{}root\textgreater{}/}group\emph{/visit1},
and per\_study data in \\ \emph{\textless{}root\textgreater{}/}group/group
(Table \ref{tbl:repo-structure}).

@@

While derivatives generated by a Study object are named in accordance
with the data specification of the Study class, input file sets and fields
do not need to be conventionally named when using basic or XNAT
repositories. In these cases, a selection stage matches input data to
entries in the data specification of the Study class. The criteria for
these selections are passed to the
the \emph{inputs} argument of the Study object at instantiation in a
dictionary mapping data specification names to \emph{FilesetInput} and
\emph{FieldInput} objects.

@@

If the data to select has been derived from an alternative Study
instance, then the name of the alternative study can be passed to the
\emph{from\_study} keyword argument provided to the Input object. There are
no restrictions on selecting any data, derived or otherwise, to match
input or derived specifications. For example, it is possible to
deliberately skip analysis steps by selecting an upstream derivative
from a previously run workflow as a match for a downstream derivative, although this is
not recommended as standard practice.

Specific file sets and fields can be passed as inputs to a Study in
\emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
specifcation). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the default
for a data specification via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).

When using BIDS repositories, the input selection stage is typically
included in the data specification (see \emph{Data and parameter
specifications)} so inputs do not need to be provided to the
initialisation of the Study. However, BidsInput and
BidsAssocInput objects can be provided to override the default
selections if required.

@@

Processor modules control how pipelines generated by a Study
are executed. There currently three Processor modules implemented in
Arcana: \emph{SingleProc}, \emph{MultiProc},
\emph{SlurmProc}, which wrap the LinearPlugin, MultiProcPlugin and SlurmPlugin
Nipype execution plugins, respectively. The main task performed
by the processor, as separate from the Nipype execution plugin it wraps,
is to determine which pipelines need to be run and link them into a
single workflow. Since this logic is implemented in the Processor
abstract base class, wrapping additional Nipype plugins as required is
trivial.

@@

Provenance is stored for each pipeline run alongside the generated
derivatives and consists of interface parameter values,
software versions of external requirements, Arcana and Nipype, 
a graph representation of the underlying Nipype workflow, checksums of inputs
and outputs, and any subject and visit IDs that were joined over in the workflow.

For subsequent analyses, changes with respect to any of the stored provenance values will
be flagged as a mismatch by default, with the exception of interface package versions
(e.g. Nipype or Banana versions). How mismatches are handled depends on the \emph{reprocess} flag passed to the
Processor. If \emph{reprocess} is true then the sessions with mismatching provenance
will be reprocessed, otherwise if \emph{reprocess} is false (the default) an exception will be raised.
Changes with respect to any element in the provenance JSON document can be ignored by providing
a list of JSONPath queries to the \emph{prov\_ignore} keyword argument.

@@

The software packages installed on the system that are required by a study's
workflows (e.g. FSL, SPM) are detected and managed by
the Environment object passed to the study at initialisation.
There are currently two types of Environment class implemented in Arcana:
\emph{StaticEnv} and \emph{ModulesEnv}.

@@

The Arcana framework is distributed as a publicly available software
package via GitHub (\url{github.com/MonashBI/arcana}) and the
Python Package Index (PyPI) (\url{pypi.org/project/arcana/}).
Study classes for T1, T2* and diffusion weighted MRI data have been
implemented as part of the \emph{Biomedical imging ANAlysis iN Arcana (Banana)} package
(\url{github.com/MonashBI/banana}; \url{pypi.org/project/banana}).
All three classes, \emph{T1Study}, \emph{T2starStudy} and
\emph{DwiStudy}, inherit generic image analysis methods, such as
registration and brain extraction, from the base class \emph{MriStudy}.

@@

The framework provided by Arcana should facilitate the development of workflows
that are consistent with F.A.I.R. guiding principles \citep{wilkinson_fair_2016},
particularly with regards to interoperability and reusability.
and their extensible, modular architecture should encourage code reuse. Every derivative created
by arcana is accompanied with sufficient provenance information to
fully reproduce the workflow. Analyses implemented in Arcana are interoperable between different
repository and execution environments. In addition, the implementation of neuroimaging
analyses in Banana is planned to closely follow the specification for BIDS
derivatives \citep{gorgolewski_bids_nodate} to enable interoperability with other tool chains.
Collaboration within a common code-base should help improve the visibility of
workflow implementations.
open source and includes text descriptions and citations of the relevant literature.

>>>>

Study classes encapsulate a study dataset (e.g. data collected across
multiple subjects using the same acquisition protocol) with the suite of
analytical methods that can be applied to the dataset. The hierarchy of
a study dataset is assumed to have two levels, \emph{subjects} and
\emph{sessions}, with each session for each subject corresponding to a
specific \emph{visit}, e.g. timepoint in longitudinal study. Derivatives
can be created at any point in this hierarchy: per-session, per-subject,
per-visit and per-study. Iteration over subjects and visits is handled
implicitly by the framework. All Study classes must inherit from the
\emph{Study} base class and be created by the \emph{StudyMetaClass}
metaclass or subclasses thereof.

@@

At the heart of each Study class is the data specification class attribute, which
specifies the input and output data of the analysis, and all
stages in between. There is a one-to-one relationship between entries in
the data specification and derivatives that are stored in the repository,
or will be stored if and when they are derived. Which intermediate
derivatives to include in the data specification, and therefore
store in the repository, is left to the discretion of the researcher
designing the analysis. However, as a general rule, derivatives that
require manual QC or are likely to be reused between different branches
of analysis should be included in the data specification.

Entries in the data specification can be of either a
\emph{file-set} or \emph{field} type. File-sets represent single files,
or sets of related files typically contained within a directory
(e.g. a multi-volume DICOM dataset). Fields
represent integer, floating point, date-time, or unicode string
values. By default, a field contains a single value but if the \emph{array}
flag is set, a field can contain a list of values.
Each file set references a \emph{FileFormat} object, which
specifies the formats of the files in the set. File formats are
explicitly registered by the researcher at design time using the
\emph{FileFormat.register(format)} class method to avoid conflicts where
the same extension is used for different formats in different contexts.

The data specification of a Study class are defined by providing a
list of \emph{FilesetSpec} and \emph{FieldSpec} objects in the
\emph{add\_data\_specs} class attribute (Figure \ref{fig:data_spec}).
The lists are collated with data specifications in any base Study
classes during the creation of the class. Input data to the study are
distinguished from derivatives by using the \emph{FilesetInputSpec}
and \emph{FieldInputSpec} subtypes. However, the distinction between
input and derived data is somewhat blurred, as derivative specification
entries can be
passed input data when the class is instantiated. In addition,
when subclassing a Study class or combining it in a MultiStudy class,
derivative specification entries can be overridden by input
specification entries, and vice-versa.

@@

All data specification entries have a \emph{frequency} attribute which specifies
where the data sits in the hierarchy of the
dataset and can take the values 'per\_session',
'per\_subject, 'per\_visit' or 'per\_study'. In
addition, derived specification entries are passed the name of a method in the
class that constructs the pipeline to derive them. Therefore, while a
pipeline can have multiple outputs, each derivative is derived by only
one pipeline.

@@

The syntax for pipeline construction is inspired by changes proposed for Nipype v2.0
(\url{github.com/nipy/nipype/issues/2539}).
Within a pipeline constructor method, a Pipeline object is constructed by the \emph{Study.new\_pipeline} method
(Figure \ref{fig:pipeline_constructor}). At initialisation, each pipeline is assigned a name,
which must be unique amongst the pipelines constructed by the Study given any combination of
switches. The methods implemented by the pipeline can be characterised by providing a the list of citations
and a text description to the \emph{citations} and \emph{desc} keyword arguments, respectively.

The \emph{add(name, interface)} method is used to add a node to a pipeline. It takes
a unique name for the node (within the pipeline) and a Nipype Interface object, and returns
a reference to the newly added node. For clarity, it is recommended to put all static inputs
(i.e. parameters) of the interface as keyword arguments of the interface constructor
(Figure \ref{fig:pipeline_constructor}). However, if an input conditionally depends on a parameter of the
study it can be set via the \emph{inputs} attribute of the returned Nipype node
(e.g. \lstinline{my_node.inputs.my_param = 1.0}).

Node inputs and ouputs are connected to each other, and to inputs and outputs of the pipeline, by
providing \emph{inputs} and \emph{outputs} keyword arguments to the \emph{add} method. Both
arguments take a dictionary. The keys of the \emph{inputs} dictionary correspond to
trait names in the node's input specification, whereas the keys of the \emph{outputs} dictionary correspond to
names of entries in the study data specification. The values of both dictionaries are 2-tuples. For pipeline
inputs, values of the \emph{inputs} dictionary consist of a name of an entry in the study data specification
and the format the input data is expected in (i.e. a FileFormat for Fileset specifications or core Python type for Field specifications).
For pipeline outputs, values of the \emph{outputs} dictionary consist of a trait name in the node's output specification and the format the
output data is generated in. For input connections from other nodes, values of the \emph{inputs} dictionary
consist of a reference to the upstream node and the name of a trait in the
upstream node's output specification (output connections to other nodes are implied by input connections to the receiving nodes).

@@

Any external software packages required by a node should be referenced
in the \emph{requirements} keyword argument as a list of
\emph{Requirement} objects when the node added to the pipeline.
Similarly, the expected memory requirements in MB and
wall time for the node execution should be provided to the keyword arguments
\emph{mem\_gb}, and \emph{wall\_time}.

@@

Study parameters can be accessed during pipeline construction with the
\emph{Study.parameter(name)} method. If conditional logic is included in
the workflow construction that alters the pipeline inputs, outputs or
parameters then it should be controlled by a switch instead of a
parameter. The analysis branch designated by a switch value should be
tested with \emph{Study.branch(switch\_name)} in the case of boolean switches and
\emph{Study.branch(switch\_name, branch\_name)} in the case of string switches.

@@

Each Study instance is assigned a name, which is used to differentiate its
results from alternative analyses on the same dataset (e.g. with different
parameterisations). Parameters are set on initialisation of the Study object
along with the range of subject and visit IDs to be included in the analysis.
If subject and visit IDs are not explicitly provided, then all IDs found in the repository are included.
The remaining arguments passed to the Study object at initialisation are the
Repository, Processor and Environment modules and a list of
\emph{Input} objects to match data in the repository with inputs in the
data specification (Figure \ref{fig:study_application}).

@@

Summary data (i.e. with \emph{per\_subject}, \emph{per\_visit}, and
\emph{per\_study} frequencies\emph{)} are stored in specially named
subjects and visits (e.g. `group'), the names for which are specified when
the repository is initialised. For example, given a basic
repository using `group' as the summary name for both subjects and visits,
per\_subject data for `subj1' would be stored at
\emph{\textless{}root\textgreater{}/subj1/}group, per\_visit data for
`visit1' in \emph{\textless{}root\textgreater{}/}group\emph{/visit1},
and per\_study data in \\ \emph{\textless{}root\textgreater{}/}group/group
(Table \ref{tbl:repo-structure}).

@@

While derivatives generated by a Study object are named in accordance
with the data specification of the Study class, input file sets and fields
do not need specific names when using basic or XNAT
repositories. In these cases, a selection stage matches input data to
entries in the data specification of the Study class. The criteria for
these selections are passed to the
the \emph{inputs} argument of the Study object at instantiation in a
dictionary mapping data specification names to \emph{FilesetInput} and
\emph{FieldInput} objects.

@@

If the data to select has been derived from an alternative Study
instance, then the name of the alternative study can be passed to the
\emph{from\_study} keyword argument provided to the Input object. There are
no restrictions on selecting any data, derived or otherwise, to match
a specification entry. For example, it is possible to
deliberately skip analysis steps by selecting an upstream derivative
from a workflow that was run previously as a match for a downstream derivative,
although this is not recommended as standard practice.

Specific file sets and fields can be passed as inputs to a Study in
\emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
specifcation). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the default
for a data specification entry via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).

When using BIDS repositories, the input selection stage is typically
included in the data specification (see \emph{Data and parameter
specifications)} so inputs do not need to be provided to the
initialisation of the Study. However, BidsInput and
BidsAssocInput objects can be provided as inputs to override the default
selections if required.

@@

Processor modules control how pipelines generated by a Study
are executed. There currently three Processor modules implemented in
Arcana: \emph{SingleProc}, \emph{MultiProc},
\emph{SlurmProc}, which wrap the LinearPlugin, MultiProcPlugin and SlurmPlugin
Nipype execution plugins, respectively. The main task performed
by the processor, as separate from the Nipype execution plugin it wraps,
is to determine which pipelines need to be run and link them into a
single workflow. Since this logic is implemented in the Processor
abstract base class, wrapping additional Nipype plugins as required is
straightforward.

@@

Provenance is stored for each pipeline run alongside the generated
derivatives and consists of the supplied node parameter values,
software versions of external requirements, Arcana and Nipype used, 
a graph representation of the underlying Nipype workflow, checksums of inputs
and outputs, and any subject and visit IDs that were joined over in the workflow.

For subsequent analyses, changes with respect to any of the stored provenance values will
be flagged as a mismatch by default, with the exception of interface package versions
(e.g. Nipype or Banana versions). How mismatches are handled depends on the \emph{reprocess} flag passed to the
Processor. If \emph{reprocess} is true then the sessions with mismatching provenance
will be reprocessed, otherwise if \emph{reprocess} is false (the default) an exception will be raised.
Changes with respect to any element in the provenance JSON document can be ignored by providing
a list of JSONPath queries to the \emph{prov\_ignore} keyword argument of the Processor.

@@

The software packages installed on the system that are required by a
workflows (e.g. FSL, SPM) are detected and managed by
the Environment object passed to the study at initialisation.
There are currently two types of Environment class implemented in Arcana:
\emph{StaticEnv} and \emph{ModulesEnv}.

@@

The Arcana framework is distributed as a publicly available software
package via GitHub (\url{github.com/MonashBI/arcana}) and the
Python Package Index (PyPI) (\url{pypi.org/project/arcana/}).
Study classes for T1, T2* and diffusion-weighted MRI data have been
implemented as part of the \emph{Biomedical imging ANAlysis iN Arcana (Banana)} package
(\url{github.com/MonashBI/banana}; \url{pypi.org/project/banana}).
All three classes, \emph{T1Study}, \emph{T2starStudy} and
\emph{DwiStudy}, inherit generic image analysis methods, such as
registration and brain extraction, from the base class \emph{MriStudy}.

@@

The framework provided by Arcana should facilitate the development of workflows
that are consistent with the F.A.I.R. guiding principles \citep{wilkinson_fair_2016}.
Each pipeline is accompanied by a text description and
a list of citations to describe the techniques it employs.
The extensibility and modularity of workflows in Arcana should encourage
their reuse, and enable communal development of code repositories.
Arcana analyses are interoperable between different
repository and execution environments. In addition, the development of
Banana is planned to closely follow the specification for BIDS
derivatives \citep{gorgolewski_bids_nodate} to enable interoperability with
other BIDS apps.
